|

### 2021-10-13 06-Chp-IPD-Analysis.Rmd
End of day

```{r, echo = FALSE, warning=FALSE, message=FALSE}
source("R/helper.R")
knitr::read_chunk("R/05-Code-IPD-Analysis.R")
doc_type <- knitr::opts_knit$get('rmarkdown.pandoc.to') # Info on knitting format
```

>>>>

```{r, echo = FALSE, warning=FALSE, message=FALSE}
source("R/helper.R")
knitr::read_chunk("R/06-Code-IPD-Analysis.R")
doc_type <- knitr::opts_knit$get('rmarkdown.pandoc.to') # Info on knitting format
```

### 2021-10-13 manuscripts/cprd/manuscript.Rmd
End of day

__Background:__ There is some evidence that circulating blood lipids play a role in the development of Alzheimer's disease (AD) and dementia. These modifiable risk factors could be targeted by existing lipid-regulating agents, including statins, for the prevention of dementia. Here, we test the association between lipid-regulating agents and risk of dementia and related conditions in the Clinical Practice Research Datalink (CPRD), a United Kingdom electronic health record database.

__Methods:__ A retrospective cohort study was performed using routinely collected data from the CPRD (January 1995 and March 2016). Multivariable Cox proportional hazard models, allowing for a time-varying treatment indicator, were used to estimate the association between seven lipid-regulating drug classes (vs. no drug) and five dementia outcomes (all-cause, vascular and other dementias, and probable and possible Alzheimer's disease).

@@

* A large cohort of patients from the Clinical Practice Research Datalink electronic health record database was assembled to examine the association of lipid-regulating agents, such as statins, with dementia outcomes.

* Little evidence that lipid-regulating agents were associated with Alzheimer’s disease, but there was some evidence for a harmful association with all-cause, vascular and other dementias. In all cases, the estimated associations were driven by the any statin subgroup, which comprised most participants in our cohort.

@@

An *a priori* protocol for this study was published,[@walker2016] and amendments to this are recorded in Supplementary Materials 1. This study was reported in line with the RECORD guidelines (Supplementary Table 1).[@nicholls2016record]

@@

Participants were included in our study cohort if their record contained any of the following index events: (a) a Read code for a diagnosis of hypercholesterolemia or related condition; (b) a Read code for prescription of a LRA (such as statins); (c) a total cholesterol test result of >4mmol/L; (d) or a low-density lipoprotein cholesterol (LDL-c) test result of >2mmol/L.

@@

The analysis was adjusted for a range of baseline covariates including sex, grouped year of entry into the cohort (<=2000, 2001-2005, 2006-2010, >2010), Charlson co-morbidity index, Index of Multiple Deprivation (IMD), consultation rate, alcohol use (current, former,  never), smoking (current, former,  never), BMI, baseline total cholesterol, and history of cardiovascular disease, coronary bypass surgery, coronary artery disease, peripheral arterial disease, hypertension, chronic kidney disease, and Type 1 and Type 2 diabetes. These variables were selected as potential confounders between dementia outcomes and use of a LRA. All covariates were determined at the index date and definitions for each can be found in Supplementary Table 2.

@@

A total of `r format(attrition[14,1], big.mark = ",")` participants met the inclusion criteria for our cohort (see Supplementary Figure 2 for the attrition flowchart), with a total follow-up of `r comma(total_followup)` patient years at risk. Most participants were included in the cohort due to an elevated test result (`r index_event_text`). The median age at the index date was `r age_text` and participants were followed up for a median of `r fu_text`. During follow-up, an all-cause dementia diagnosis was recorded for `r format(p1[1,2], big.mark = ",")` patients  (`r format(p1[2,2], big.mark = ",")` probable Alzheimer's disease, `r format(p1[3,2], big.mark = ",")` possible Alzheimer's disease, `r format(p1[4,2], big.mark = ",")` vascular dementia, `r format(p1[5,2], big.mark = ",")` other dementia). The number of events, time-at-risk and crude rates for each drug class, tabulated by dementia outcome, are shown in Table \@ref(tab:followUp-table). Most participants (`r percentage.statins`) prescribed a lipid-regulating agent were prescribed a statin. We excluded the "Ezetimibe and statins" and "Nicotinic acid groups" classes from subsequent subgroup analyses based on the extremely small number of participants in these groups (Table \@ref(tab:followUp-table)). The distribution of baseline characteristics across the remaining seven drug classes can be seen in Table \@ref(tab:cprdCharacteristics-table). The stopping, addition and switching of drug classes was common across all exposure groups (Supplementary Table 3).

@@

In contrast to the findings for Alzheimer's disease outcomes, lipid-regulating agents were associated with an increased risk of a subsequent diagnosis of vascular dementia (`r vasdem_text`) or other dementias (`r othdem_text`). Again, the observed estimate was driven mainly by the any statin subgroup, but there was some evidence that ezetimibe was associated with an increased risk of vascular (`r vasdem_eze_text`) and other (`r othdem_eze_text`) dementia.

@@

Much of the existing literature focuses on the association of statins alone with neurodegenerative outcomes, with other lipid-regulating agents being grouped as "non-statin cholesterol-lowering drugs".[@ancelin2012] echoing the distribution of participants in our analysis.

@@

Our additional analyses of statin properties found little evidence of differences in associations of lipophilic and hydrophilic statins and incidence of Alzheimer’s disease, consistent with a recent meta-analysis of observational studies.[@chu2018]

@@

However, the findings of our analysis are subject to several limitations. Confounding by indication is a major bias in pharmacoepidemiological studies and could provide a potential explanation for the observed increased risk of vascular and other dementias with lipid regulating agent use. Patients with vascular risk factors are more likely to receive a statin prescription and also to be diagnosed with vascular dementia. Supporting evidence for this interpretation comes from a variety of sources, including the results of the control outcome analyses. This would explain the increased risk for the ischaemic heart disease outcome, for which statins are known to be protective, whilst almost no association was observed with the backpain outcome, indicating that most of the uncontrolled confounding is likely to be related to vascular factors. Additionally, we obtained the expected harmful result for Type 2 diabetes, where statins' mechanism of action on this outcome is unlikely to be vascular,[@collins2016a; @macedo2014; @smit2020] Further supporting evidence comes from the increasingly harmful association observed when moving from the probable and possible Alzheimer's disease outcome to the other dementias outcome, and finally to the vascular dementia outcomes. This pattern suggests that the strength of the residual confounding by indication increases as the proportion of cases with a vascular component in an outcome definition increases. A review of other available literature suggests that this observation (a harmful effect of lipid regulating agents on vascular-related outcome due to confounding by indication) is not unusual. Using a conventional epidemiological technique, a previous analysis also found an increased risk of coronary heart disease (analogous to the ischaemic heart disease outcome used in our analysis) in those taking statins (`r estimate(1.31,1.04, 1.66, type = "HR")`).[@danaei2013b] Control for confounding by indication in that study through the use of a trial emulation analysis gave an estimate of 0.89 (95%CI: 0.73-1.09), a comparable though less precise estimate to that observed in RCTs of statin use (`r estimate(0.73,0.67,0.80,type="")`).[@taylor2013]

A secondary limitation is the potential for differential outcome misclassification based on the exposure, as we cannot exclude the possibility that for people with memory complaints, a diagnosis of vascular dementia might be made more frequently than Alzheimer’s disease if their medical records contain prescriptions for lipid-regulating agents. Further, there is also the potential for non-differential misclassification of the outcome ,based on the use of electronic health records to identify dementia cases.[@wilkinson2018; @mcguinness2019validity]

@@

This is most likely to explain the unexpected increase in risk of vascular dementia associated with statin use. Future research should aim to address these biases using newer methods such as a trial emulation analysis.

@@

__Data/code availability__
This analysis used the CPRD-GOLD primary care dataset March 2016 snapshot (ISAC 15_246R), which is available upon application to the CPRD Independent Scientific Advisory Committee. The code lists used to define the outcomes and covariates for this study, in addition to the cleaning and analysis scripts used to create the study cohort and perform the analyses, are available on GitHub (https://github.com/mcguinlu/CPRD-LRA), and were archived at the time of publication on Zenodo (DOI: __TBC__).

@@

__Conflict of interest statement__
None declared.

@@

# Tables
<!----------------------------------------------------------------------------->
(ref:followUp-caption) Summary of person-time at risk and number of events by exposure group.

>>>>

__Background:__ There is some evidence that circulating blood lipids play a role in the development of Alzheimer's disease (AD) and dementia. These modifiable risk factors could be targeted by existing lipid-regulating agents, including statins, for dementia prevention. Here, we test the association between lipid-regulating agents and incidence of dementia and related conditions in the Clinical Practice Research Datalink (CPRD), an United Kingdom-based electronic health record database.

__Methods:__ A retrospective cohort study was performed using routinely collected CPRD data (January 1995 and March 2016). Multivariable Cox proportional hazard models, allowing for a time-varying treatment indicator, were used to estimate the association between seven lipid-regulating drug classes (vs. no drug) and five dementia outcomes (all-cause, vascular and other dementias, and probable and possible Alzheimer's disease).

@@

* A large cohort of patients from the Clinical Practice Research Datalink (CPRD) electronic health record database was assembled to examine the association of lipid-regulating agents, such as statins, with dementia outcomes.

* There was little evidence that lipid-regulating agents were associated with Alzheimer’s disease, but there was some evidence for a harmful association with all-cause, vascular and other dementias. In all cases, the estimated associations were driven by the any statin subgroup, which comprised most participants in our cohort.

@@

An *a priori* protocol for this study was published,[@walker2016] and amendments to this are recorded in Supplementary Materials 1. This study was reported in line with the RECORD guidelines (Supplementary Materials 2).[@nicholls2016record]

@@

Participants were included in our study cohort if their record contained any of the following index events: (a) a code for a diagnosis of hypercholesterolemia or related condition; (b) a code for prescription of a LRA (such as statins); (c) a total cholesterol test result of >4mmol/L; (d) or a low-density lipoprotein cholesterol (LDL-c) test result of >2mmol/L.

@@

The analysis was adjusted for a range of baseline covariates including sex, grouped year of entry into the cohort (<=2000, 2001-2005, 2006-2010, >2010), Charlson co-morbidity index, Index of Multiple Deprivation (IMD), consultation rate, alcohol use (current, former,  never), smoking (current, former,  never), BMI, baseline total cholesterol, and history of cardiovascular disease, coronary bypass surgery, coronary artery disease, peripheral arterial disease, hypertension, chronic kidney disease, and Type 1 and Type 2 diabetes. These variables were selected as potential confounders between dementia outcomes and use of a LRA. All covariates were determined at the index date and definitions for each can be found in Supplementary Table 1.

@@

A total of `r format(attrition[14,1], big.mark = ",")` participants met the inclusion criteria for our cohort (see Supplementary Figure 2 for the attrition flowchart), with a total follow-up of `r comma(total_followup)` patient years at risk. Most participants were included in the cohort due to an elevated test result (`r index_event_text`). The median age at the index date was `r age_text` and participants were followed up for a median of `r fu_text`. During follow-up, an all-cause dementia diagnosis was recorded for `r format(p1[1,2], big.mark = ",")` patients  (`r format(p1[2,2], big.mark = ",")` probable Alzheimer's disease, `r format(p1[3,2], big.mark = ",")` possible Alzheimer's disease, `r format(p1[4,2], big.mark = ",")` vascular dementia, `r format(p1[5,2], big.mark = ",")` other dementia).

The number of events, time-at-risk and crude rates for each drug class, tabulated by dementia outcome, are shown in Table \@ref(tab:followUp-table). Most participants (`r percentage.statins`) prescribed a lipid-regulating agent were prescribed a statin. We excluded the "Ezetimibe and statins" and "Nicotinic acid groups" drug classes from subsequent subgroup analyses based on the extremely small number of participants in these groups (Table \@ref(tab:followUp-table)). The distribution of baseline characteristics across the remaining seven drug classes can be seen in Table \@ref(tab:cprdCharacteristics-table). The stopping, addition and switching of drug classes was common across all exposure groups (Supplementary Table 2).

@@

In contrast to the findings for Alzheimer's disease outcomes, lipid-regulating agents were associated with an increased risk of a subsequent diagnosis of vascular dementia (`r vasdem_text`) or other dementias (`r othdem_text`). Again, the observed estimate was driven mainly by the any statin subgroup, but there was some evidence that ezetimibe was associated with an increased risk of vascular (`r vasdem_eze_text`) and other (`r othdem_eze_text`) dementias.

@@

Much of the existing literature focuses on the association of statins alone with neurodegenerative outcomes, with other lipid-regulating agents being grouped as "non-statin cholesterol-lowering drugs",[@ancelin2012] echoing the distribution of participants in our analysis.

@@

Our additional analyses stratified by statin properties found little evidence of differences in associations of lipophilic and hydrophilic statins and incidence of Alzheimer’s disease, consistent with a recent meta-analysis of observational studies.[@chu2018]

@@

However, the findings of our analysis are subject to several limitations. Confounding by indication is a major bias in pharmacoepidemiological studies and could provide a potential explanation for the observed increased risk of vascular and other dementias with lipid regulating agent use. Patients with vascular risk factors are more likely to receive a statin prescription and also to be diagnosed with vascular dementia. Supporting evidence for this interpretation comes from a variety of sources, including the results of the control outcome analyses. This would explain the increased risk for the ischaemic heart disease outcome, for which statins are known to be protective, whilst almost no association was observed with the backpain outcome, indicating that most of the uncontrolled confounding is likely to be related to vascular factors. Additionally, we obtained the expected harmful result for Type 2 diabetes, where statins' mechanism of action on this outcome is unlikely to be vascular.[@collins2016a; @macedo2014; @smit2020] Further supporting evidence comes from the increasingly harmful association observed when moving from the probable and possible Alzheimer's disease outcome to the other dementias outcome, and finally to the vascular dementia outcomes. This pattern suggests that the strength of the residual confounding by indication increases as the proportion of cases with a vascular component in an outcome definition increases. A review of other available literature suggests that this observation (a harmful effect of lipid regulating agents on vascular-related outcome due to confounding by indication) is not unusual. Using a conventional epidemiological technique, a previous analysis also found an increased risk of coronary heart disease (analogous to the ischaemic heart disease outcome used in our analysis) in those taking statins (`r estimate(1.31,1.04, 1.66, type = "HR")`).[@danaei2013b] Controlling for confounding by indication in that study through the use of a trial emulation analysis gave an estimate of 0.89 (95%CI: 0.73-1.09), a comparable though less precise estimate to that observed in RCTs of statin use (`r estimate(0.73,0.67,0.80,type="")`).[@taylor2013]

A secondary limitation is the potential for differential outcome misclassification based on the exposure, as we cannot exclude the possibility that for people with memory complaints, a diagnosis of vascular dementia might be made more frequently than Alzheimer’s disease if their medical records contain prescriptions for lipid-regulating agents. Further, there is also the potential for non-differential misclassification of the outcome, based on the use of electronic health records to identify dementia cases.[@wilkinson2018; @mcguinness2019validity]

@@

This is likely to explain the unexpected increase in risk of vascular dementia associated with statin use. Future research should aim to address these biases using newer methods such as a trial emulation analysis.

@@

__Data/code availability__

This analysis used the CPRD-GOLD primary care dataset March 2016 snapshot (ISAC 15_246R), which is available upon application to the CPRD Independent Scientific Advisory Committee. The code lists used to define the outcomes and covariates for this study, in addition to the cleaning and analysis scripts used to create the study cohort and perform the analyses, are available on GitHub (https://github.com/mcguinlu/CPRD-LRA), and were archived at the time of publication on Zenodo (DOI: __TBC__).

@@

__Conflict of interest statement__

None declared.

@@

# Tables
<!----------------------------------------------------------------------------->
(ref:followUp-caption) Summary of number of events, participant-time-at-risk and crude rates by drug class and dementia outcome.

### 2021-10-13 manuscripts/cprd/supp-materials.Rmd
End of day

## Supplementary Table 1: RECORD Reporting Guidelines {#supp-tab-strobe}

@@

## Supplementary Table 2: Definition of exposures and covariates

@@

## Supplementary Table 3: Adherence and switching by drug class

>>>>

## Supplementary Material 2: RECORD Reporting Guidelines {#supp-tab-strobe}

@@

## Supplementary Table 1: Definition of exposures and covariates

@@

## Supplementary Table 2: Adherence and switching by drug class
